---
title: "Seattle Crime Analysis by Dates and Times"
author: "John Fitzpatrick"
date: "January 10, 2016"
output: html_document
---

# Analysis of Crime frequency in Seattle 
This analysis will attempt to highlight the frequency of crimes in the summer of
2014 for the city of Seattle. The analysis will look at crime counts 
* Date
* Day of the week
* Hour of the Day

It's hoped that this analysis would aid police planners in allocating manpower
and other resources to the times when crimes are most frequently committed


```{r read_and_format_data,echo=FALSE,message=FALSE}
library(ggplot2)
library(dplyr)
library(lubridate)

train_ds_path <- "..\\data\\seattle_incidents_summer_2014.csv"

col_classes <- c(
"character"
,"character"
,"factor"
,"character"
,"factor"
,"factor"
,"character"
,"character"
,"character"
,"character"
,"character"
,"factor"
,"factor"
,"character"
,"numeric"
,"numeric"
,"character"
,"integer"
,"integer"
)

dt <- read.table(train_ds_path
                 ,header = TRUE
                 ,sep = ","
                 ,quote = "\"'"
                 ,stringsAsFactors = FALSE
                 ,colClasses = col_classes
)

colnames(dt)[9] <- "Occurred.Date.Range.Start"

dt$Date.Reported <- mdy_hms(dt$Date.Reported, tz = "America/Los_Angeles")

dt$Occurred.Date.Range.Start <- mdy_hms(dt$Occurred.Date.Range.Start, tz = "America/Los_Angeles")
dt$Occurred.Date.Range.End <- mdy_hms(dt$Occurred.Date.Range.End, tz = "America/Los_Angeles")
dt$Occurred.Date <- as.Date(dt$Occurred.Date.Range.Start)
dt$wday <- wday(dt$Occurred.Date.Range.Start, label = TRUE)
dt$hour <- hour(dt$Occurred.Date.Range.Start)

```

### Crimes by Date
```{r crimes_by_date,echo=FALSE}
by_date <- group_by(dt, Occurred.Date)
crimes_by_dt <- summarise(by_date, count=length(General.Offense.Number))
p1 <- ggplot(data = crimes_by_dt, aes(x = Occurred.Date, y = count)) +
  xlab("Date") +
  ylab("Count") +  
  ggtitle("Seattle Crime Counts (Summer 2014)") +
  geom_smooth(method="lm") +
  geom_line() 

print(p1)
```

We can see that there's a lot of variability in the crime counts 
from day to day. But that there's a general downward trend from the
beginning of the summer to the end of the summer. Without further
information I believe the drop in crimes on the last day is likely
due to some type of reporting anomaly since it's over twice as low
as the next low crime day

```{r explain_low_count,echo=FALSE}
head(crimes_by_dt[order(crimes_by_dt$count),],5)

```

### Crimes by Day of Week
```{r crimes_by_wday,echo=FALSE}
p2 <- qplot(dt$wday,
            geom="bar",
            main = "Seattle Crime Counts (Summer 2014)", 
            xlab = "Day of Week",
            ylab = "Count",              
            fill=I("deepskyblue1"), 
            col=I("red")
)
print(p2)

```

There seems to be a noticeable increase in crime on Fridays compared to the 
other days in the week

### Crimes by Hour of the Day
```{r crimes_by_hour,echo=FALSE}
p3 <- qplot(dt$hour,
            geom="bar",
            main = "Seattle Crime Counts (Summer 2014)", 
            xlab = "Hour of Day",  
            ylab = "Count",                          
            fill=I("deepskyblue1"), 
            col=I("red"),
)

print(p3)
```

Crime is highest between midnight and 1AM and then drops sharply during the wee
hours of the morning. It then starts picking up again around 6AM and increases steadily with a peak around noon. It remains high for the rest of the day until
the new after-midnight peak is reached.
